Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia

### 2024-11-05 analysis/paper/paper.qmd
- Small grammar improvements (thank you Jack and Cedric)

---
title: "Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
author:
  - Joana Belmiro:
      orcid: https://orcid.org/0000-0003-4951-2907
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute: 
        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
  - Salvador Dominguez-Bella:
      orcid: https://orcid.org/0000-0003-3892-763X
      institute: 
        - UGEA-PHAM, Departamento de Ciencias de la Tierra, Facultad de Ciencias, Universidad de Cádiz. Puerto Real, Cádiz, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated" 
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/springer-socpsych-author-date.csl" # Insert path for the bib-style
abstract: |
  Analyses of raw materials and the distinction between local/regional and long-distance sources have proven invaluable for understanding the extensive movements, interactions, and social networks during the Upper Palaeolithic in the Iberian Peninsula. However, unlike other parts of Iberia, research on the management and acquisition of raw materials in the south and west of Iberia remains relatively underdeveloped. Despite significant knowledge about the technological practices of Palaeolithic hunter-gatherers from Southern Portugal, particularly from studies conducted at the site of Vale Boi, there is a noticeable lack of focus on raw materials management. This paper presents the first comprehensive characterization of chert raw materials from the Gravettian, Proto-Solutrean, and Solutrean occupations at Vale Boi, using both macroscopic and petrographic techniques. Our study reveals that the majority of chert found at Vale Boi originates locally, within a 20 km radius. However, a non-neglectable portion of the chert comes from non-local sources, indicating >200 km raw material circulation from Central Portugal and Southern Spain.
keywords: |
  Upper Paleolithic; Iberian Peninsula; Lithic raw materials; Petrography
---

@@

Based on the notions that the role of Vale Boi during the Solutrean as a connection point between different regions may have promoted the exchange of information and culture, including raw materials from long-distances, and the limited circulation of people during the Gravettian may have isolated groups occupying Vale Boi, making them mostly reliant on the available local or regional resources, a set of hypotheses and expectations regarding chert procurement and human mobility throughout the UP at Vale Boi can be proposed. On one hand, it would be expected that Solutrean occupations are characterized by the predominant use of local raw materials but with a considerable percentage of cherts obtained from long-distance sources, possibly through exchange with different groups in Iberia. Conversely, it would be expected that Gravettian occupations would be mostly composed of local cherts, with a more limited presence of long-distance cherts.

@@

The availability of regional resources containing cherts and the characterization of their lithologies has already been published by our team [@belmiroCreatingFramesReference2023], and we herein present the summarized main findings. The results showed the existence of abundant chert nodules and beds found in the Lower Jurassic (Carixian) formation and Upper Jurassic (Kimmeridgian) formation, mostly outcropping in the cliff and beach areas of Westernmost Algarve, at \~20 km southwest from the site of Vale Boi, and with one single inland outcrop located \~8 km east from the site (@fig-map b). The Lower Jurassic nodules and beds can be found in abundant outcrops, but also frequently in sub-primary or secondary deposition settings in proximity to the outcrops possibly due to natural erosion. However, in the case of the inland outcrop of Ferrel, the outcrops is partially destroyed due to recent human action, and chert can be found scattered as chunks or larger nodules. The Upper Jurassic nodules can be found sub-primary deposition and of easy access, or in secondary deposition. These cherts are frequently opaque with yellow, grey, purple, or red colours, occasionally with bands or laminations, and with frequently visible sponge spicules. Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being \~20 cm, and frequently characterized by high levels of dolomitization and fractures. Petrographically, they are composed mainly of microcrystalline quartz, with wackstone packstone textures, small percentages of macrocrystalline quartz, fibrous chalcedony and allochems composed of oxides and poorly preserved bioclasts. When identifiable, the bioclasts are Echinoderms, Radiolarians, Sponge spicules and bivalve shells. The Upper Jurassic cherts of Mareta show similar petrographic characteristics, but with abundant Calcispheres which are not always apparent under the stereomicroscope.

Chert sources were also identified in two formations from the eastern area of the Algarve (@fig-map c). The Middle Jurassic Malhão formation chert outcrops were identified at around 80-100 km east of the site and characterized by chert nodules with similar macroscopic features to those found in Western Algarve. Petrographically they are composed mainly by microcrystalline quartz and characterized by wackestone textures, while dolomite, chalcedony and macrocrystalline quartz are also present in small percentages (\<10%). Allochems are iron oxides, and albeit poorly preserved, several bioclasts can be identified: Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres and possibly Tentaculites. The Upper Jurassic Jordana formation chert outcrops were identified at around 90 km east of Vale Boi and characterized by chert nodules with significantly different macroscopic features to those of other formations, with frequently small sizes (\~5 cm), difficult removal from the parent rock and unidentified secondary deposits. Petrographically they are composed mostly of microcrystalline quartz, with wackstone to packstone textures, and small percentages of fibrous chalcedony and dolomite (\<10%). They show frequent iron oxides and very frequent bioclasts, albeit frequently poorly preserved and unidentifiable; identified fossils are fossils are Calcispheres, Bivalve shells, Sponge spicules, Ostracods, Echinoderms and Gastropods.

Chalcedony outcrops \~10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013], although previous studies also note the presence of chalcedony (with frequent presence of cortex) in the immediate vicinity of the site [@pereira_territory_2016-1; @gibajabaoProvenienceTechnologyMorphology2013]. The chalcedony frequently has fractures and fissures and is characterized by a fibrous quartz structure, without inclusions or fossils.

@@

Chert was chosen for this study as it is one of the three main used raw materials at the archaeological site of Vale Boi alongside quartz and greywacke. Other raw materials such as chalcedony, dolerite, and schists have also been identified [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @pereira_territory_2016-1].

@@

Although systematic raw material research has not been conducted on these materials, according to the regional geological context, they have been interpreted as local and readily accessible in the vicinity of the site [@pereira_territory_2016-1]. Frequently, the quartz and greywacke assemblages include large amounts of shattered and unknapped chunks or slabs, and have been interpreted in great part as functionally specialised in activities related to the fragmentation of bones in the case of quartz and greywack [@cascalheira_2017], or in grease rendering in the case of quartz [@manneIntensiveSubsistencePractices2012]. Previous studies from the Terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiroProtoSolutreanLithicTechnology2020]. Based on this, despite the smaller amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits. The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site, such as a Vale Comprido point, the fossil-director for the Proto-Solutrean in Western Portugal [@belmiro_gravettian-solutrean_2021].

@@

While the advantages of employing a macroscopic methodology for analysing the cherts of Vale Boi are notable, a limitation of this approach is its inability to unequivocally characterize and categorize artefacts that exhibit surface alterations typical of an open-air site, including those related to fire, since weathered and altered surface may destroy visible structures and even alter the chemical composition [@delluniversitaDevelopmentMultiparametricCharacterisation2019]. Other studies frequently exclude samples that display intense fire damage or several surface alterations [@delluniversitaDevelopmentMultiparametricCharacterisation2019; @gomezdesolerPanadellaChertMontmaneu2020; @sotoProcurementMobilityLate2016]. For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), and given the unexistence of a reference collection for local altered cherts, they were analysed using the individual sample dataset (Online Resource 1) but grouped in a category of indeterminate cherts (INDET). This category included key alterations such as heat/fire related alterations which clearly impacted the macroscopic characteristics of the samples, fully changing its color, luster and texture, and covering them in crazings (e.g., @fig-terrace-alterations a). Similarly, it also included alteration rinds that extended through most of the sample, impeding the identification of its original color, texture, translucency, inclusions and fossils (e.g., @fig-terrace-alterations b) and the presence of extended pitting. However, whenever surface alterations were present in only a limited area of the sample leaving recognizable macroscopic features and fossils, or when key elements were distinguishable to make a characterization (e.g., reddening in the Type 2 cherts, Online Resource 2, figure s5), the samples were not collapsed into the INDET category.

@@

```{r}

#| label: tbl-type-source
#| tbl-cap: Chert types identified during the analysis, and potential sources. The distance column refers to the distance of the potential source from Vale Boi, as the crow flies. Only the TL samples with possible identified sources were included in the table.

@@

Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiroCreatingFramesReference2023], we interpret Types 2, 3, 4, and 5 to be local, as they are congruent with the described characteristics representative of cherts from the Lower Jurassic found in the known outcrops Cabo S. Vicente, Foz dos Fornos, Ponta dos Altos, Praia do Belixe and Ferrel (Carixian; \<20 km from Vale Boi, as the crow flies), possibly Upper Jurassic Kimmeridgian outcrops of Mareta (\~15 km from Vale Boi, as the crow flies) and Middle Jurassic Malhão formations (80-100 km East of Vale Boi, as the crow flies) of the Algarve basin. Types 2 and 3 are congruent with the frequently available brown/yellow and red nodules and beds (@fig-local-types b-d), ranging from uncommon to very frequent fossils. These can be found in all previously mentioned outcrops. The macroscopic similarities of Types 4 and 5 are more limited (@fig-local-types e-h), as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the inland Lower Jurassic outcrop of Ferrel (@fig-map b). As such, we interpret these cherts as local, but possibly from nodules and inland outcrops that are no longer frequently available or visible in the landscape. This is especially relevant since the inland outcrops of Ferrel are currently mostly destroyed and can be found in poor preservation conditions, due to recent human intevention and construction work, and with small, broken nodules in the surface, frequently altered [@belmiroCreatingFramesReference2023].

@@

Studies of Upper Cretaceous cherts located in the Lisbon area describe cherts with similar macroscopic characteristics (@fig-map a), with micro-cryptocrystalline quartz mudstones with bioclasts such as sponge spicules or bivalve fragments, or wackestone/packstone textures with abundant bioclasts, including ostracods and different types of foraminifers [@jordaoFlintSourcesMobility2022]. These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from Central Portugal hosted at LusoLit. Alike the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matiasRawMaterialSourcing2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%). From these Cretaceous cherts from Central Portugal, only the dark grey nodules (sample RT231, with petrographic descriptions available in our online research compendium) showed identifiable fossils which were rare ostracods.

@@

The consultation of the lithotheque hosted at UCA allowed us to identify a subset of cherts with similarities to those from Type 7 from the Middle Subbetic region (Betic Systems, south Spain, @fig-map). These cherts are massive, sub-translucent, peloidal/oolitic packstones. Allochems include very frequent peloids and ooids. The peloids are densely arranged in the samples, more or less visible depending on the alterations to the surface or the thickness. The ooids are poorly sorted, with oval or round shapes, and replaced by quartz. Uncommonly the ooids show a preserved nucleus. The presence of bioclasts varies between samples, ranging between common to rare, and when present/identifiable include sponge spicules (common), foraminifera (rare to uncommon), and echinoderm spines (rare). Similar cherts have also been identified and described in previous works, describing grey cherts with peloidal and oolitic facies and occasional foraminifera [@rodriguezPrehistoricFlintExploitations2011]. Despite the nonexistence of thin sections for comparison, the macroscopic similarities between the cherts allow us to interpret Type 7 as coming from the Upper Jurassic formations of the Betic Systems, located in the south and southeastern Iberian Peninsula, at least 250 km (as the crow flies) east from Vale Boi (@fig-map).

Type 8 (@fig-distance-types, e-f) corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment. It represents \~2% of the overall Terrace cherts and \~5% of the chert in the Shelter. Identified and well-preserved fossils are foraminifera from the Pfenderenidae family (@fig-distance-types e, f2), which are larger benthic foraminifera from marine environments [@boudagher-fadelChapterMesozoicLarger2008]. The association of these foraminifera indicates this chert formed from sediments deposited possibly between the Jurassic to the Cretaceous. Macroscopically, this type is characterized by grey and white colours (@fig-distance-types e-f1). The translucency ranges from sub-translucent to opaque. Whenever present, the cortex is rounded and thin (@tbl-cortex_tab). Petrographically, this type is a packstone (mudstone texture close to the edges) with a massive microstructure. It is composed of 97% micro-cryptocrystalline quartz, macrocrystalline quartz, and fibrous chalcedony, found mostly replacing fossils. Other minerals include uncommon iron oxides and the porosity is of vug type (3%). Fossil content is very frequent (@fig-distance-types f2-f3), albeit mostly unidentifiable and poorly preserved. Identified fossils are rare sponge spicules, rare ostracods with some degree of preservation, common small, unidentified foraminifera (100 μm) of differing preservation degrees, and as mentioned above, foraminifera from the Pfenderenidae family (500 μm). Despite the macroscopic similarities between type 8 and cherts from the Middle Subbetic region chert formations (@fig-map), the geological reference samples showed no macroscopically visible foraminifera. However, our results indicate that despite no foraminifera being visible macroscopically, they are present in thin section. The fact that no foraminifera were identified in the small number of analysed samples from the UCA lithotheque with similarities to Type 8 (n=3) may simply reflect their uncommon visibility, possibly due to patination. Another possible source for Type 8 cherts may be the Upper Cretaceous wackestone/packstone cherts with foraminifers from the Lisbon region [@jordaoFlintSourcesMobility2022]. However, these studies do not specify the existence of Pfenderenidae foraminifers. Given the lack of this specific fossil in the reference samples from the Betic chert \~250 km east and southeast, and the Central Portugal cherts 215-250 km north, the probable source of Type 8 remains unknown.

Type 9 (@fig-distance-types g) corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone. It corresponds to less than \~2% of the overall cherts in the Terrace and Shelter areas. It is opaque, has a heterogeneous structure and is characterized by a variable colour distribution: it shows a horizontal laminated and finely laminated pattern, with grey dark bands intercalated with light grey bands (@fig-distance-types g1). Whenever present, the cortex is rounded and is generally thin (@tbl-cortex_tab). In specific samples, the cortex is present on two parallel planes of the sample (parallel to the laminations), which indicates the chert was originally available in bedded layers. Petrographically this chert is a banded packstone (banding of sedimentary origin; @fig-distance-types g2), composed mostly of micro-cryptocrystalline quartz (48%), micrite (40%), dolomite (10%), and accessory fibrous chalcedony and mica (muscovite). The allochems are peloids, unidentifiable bioclasts, iron oxides, and opaque minerals. Porosity is low (1%). Although its origin is unknown, when visiting the UCA lithotheque, we identified a small type of banded black cherts without noticeable fossils or inclusions. Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Type 9 belongs to Betic chert formations in Southern Spain, 250 km east (as the crow flies) from the site (@fig-map).

@@

count_plot <- ggplot(count_table, aes(x=Level,y=percent, fill = RMTYPE)) +
  geom_bar(stat = "identity", width = 0.5) +
  facet_grid(.~Technocomplex, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format()) +
  #theme_minimal() + #type of theme
  labs(y = "Count %") +# change labels
  guides(fill=guide_legend(title=NULL)) + # remove legend title
  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
                position=position_stack(vjust=0.5), colour="white", size=3) +
  theme(axis.title.x = element_blank()) +
  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
                              "#FACE7F", "#845460", "#EAD3CB", "#2B4F60",
                              "#996666", "#BDD2B6", "#798777", "#BDC7C9"))

@@

```{r}
#| label: chisq
#| include: false

library(corrplot)

chisq <- phase_dataset %>% 
  filter(RMTYPE != "INDET") %>% 
  filter(RMTYPE != "TYPE 11")

chisq_tab <- table(chisq$RMTYPE, chisq$Level)
chisq_matrix <- as.table(as.matrix(chisq_tab))
chisq_rm <- chisq.test(chisq_tab)
round(chisq_rm$residuals, 3)

corrplot(chisq_rm$residuals, method = "color", is.corr = FALSE)

>>>>

---
title: "Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia"
author:
  - Joana Belmiro:
      orcid: https://orcid.org/0000-0003-4951-2907
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute: 
        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
  - Salvador Dominguez-Bella:
      orcid: https://orcid.org/0000-0003-3892-763X
      institute: 
        - UGEA-PHAM, Departamento de Ciencias de la Tierra, Facultad de Ciencias, Universidad de Cádiz. Puerto Real, Cádiz, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated" 
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/springer-socpsych-author-date.csl" # Insert path for the bib-style
abstract: |
  Analyses of raw materials and the distinction between local/regional and long-distance sources have proven invaluable for understanding the extensive movements, interactions, and social networks during the Upper Palaeolithic in the Iberian Peninsula. However, unlike other parts of Iberia, research on the management and acquisition of raw materials in the south and west of Iberia remains relatively underdeveloped. Despite significant knowledge about the technological practices of Palaeolithic hunter-gatherers from Southern Portugal, particularly from studies conducted at the site of Vale Boi, there is a noticeable lack of focus on raw materials management. This paper presents the first comprehensive characterization of chert raw materials from the Gravettian, Proto-Solutrean, and Solutrean occupations at Vale Boi, using both macroscopic and petrographic techniques. Our study reveals that the majority of chert found at Vale Boi originates locally, within a 20 km radius. However, a non-negligible portion of the chert comes from non-local sources, indicating >200 km raw material circulation from Central Portugal and Southern Spain.
keywords: |
  Upper Paleolithic; Iberian Peninsula; Lithic raw materials; Petrography
---

@@

Based on the notions that a) during the Solutrean Vale Boi served as a connection point between different regions which may have promoted the exchange of information and culture, including raw materials from long-distances; and b) the limited circulation of people during the Gravettian may have isolated groups occupying Vale Boi, making them mostly reliant on the available local or regional resources, we propose a set of hypotheses and expectations regarding chert procurement and human mobility throughout the UP at Vale Boi. On one hand, it would be expected that Solutrean occupations are characterized by the predominant use of local raw materials but with a considerable percentage of cherts obtained from long-distance sources, possibly through exchange with different groups in Iberia. Conversely, it would be expected that Gravettian occupations would be mostly composed of local cherts, with a more limited presence of long-distance cherts.

@@

The availability of regional resources containing cherts and the characterization of their lithologies has already been published by our team [@belmiroCreatingFramesReference2023], and we herein present the summarized main findings. The results showed the existence of abundant chert nodules and beds found in the Lower Jurassic (Carixian) formation and Upper Jurassic (Kimmeridgian) formation, mostly outcropping in the cliff and beach areas of Westernmost Algarve, at \~20 km southwest from the site of Vale Boi, and with one single inland outcrop located \~8 km east from the site (@fig-map b). The Lower Jurassic nodules and beds can be found in abundant outcrops, but also frequently in sub-primary or secondary deposition settings in proximity to the outcrops, possibly due to natural erosion. However, in the case of the inland outcrop of Ferrel, the outcrop is partially destroyed due to recent human action, and chert can be found scattered as chunks or larger nodules. The Upper Jurassic nodules can be found in sub-primary deposition and are easily accessible, or in secondary deposition. These cherts are frequently opaque with yellow, grey, purple or red colours, occasionally with bands or laminations, and frequently with visible sponge spicules. Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being \~20 cm, and frequently characterized by high levels of dolomitization and fractures. Petrographically, they are composed mainly of microcrystalline quartz, with wackestone packstone textures, small percentages of macrocrystalline quartz, fibrous chalcedony and allochems composed of oxides and poorly preserved bioclasts. When identifiable, the bioclasts are Echinoderms, Radiolarians, Sponge spicules and bivalve shells. The Upper Jurassic cherts of Mareta show similar petrographic characteristics, but with abundant Calcispheres which are not always apparent under the stereomicroscope.

Chert sources were also identified in two formations from the eastern area of the Algarve (@fig-map c). The Middle Jurassic Malhão formation chert outcrops were identified at around 80-100 km east of the site and characterized by chert nodules with similar macroscopic features to those found in Western Algarve. Petrographically they are composed mainly by microcrystalline quartz and characterized by wackestone textures, while dolomite, chalcedony and macrocrystalline quartz are also present in small percentages (\<10%). Allochems are iron oxides, and albeit poorly preserved, several bioclasts can be identified: Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres and possibly Tentaculites. The Upper Jurassic Jordana formation chert outcrops were identified at around 90 km east of Vale Boi and characterized by chert nodules with significantly different macroscopic features to those of other formations, with frequently small sizes (\~5 cm), difficult removal from the parent rock and unidentified secondary deposits. Petrographically they are composed mostly of microcrystalline quartz, with wackstone to packstone textures, and small percentages of fibrous chalcedony and dolomite (\<10%). They show frequent iron oxides and very frequent bioclasts, albeit often poorly preserved and unidentifiable; identified fossils are Calcispheres, Bivalve shells, Sponge spicules, Ostracods, Echinoderms and Gastropods.

Chalcedony can be found \~10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013]. Previous studies also note the presence of chalcedony (with frequent presence of cortex) in the immediate vicinity of the site [@pereira_territory_2016-1; @gibajabaoProvenienceTechnologyMorphology2013]. The chalcedony frequently has fractures and fissures and is characterized by a fibrous quartz structure, without inclusions or fossils.

@@

Chert was chosen for this study as it is one of the three main raw materials used at the archaeological site of Vale Boi alongside quartz and greywacke. Other raw materials such as chalcedony, dolerite, and schists have also been identified [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @pereira_territory_2016-1].

@@

Although systematic raw material research has not been conducted on these materials, according to the regional geological context, they have been interpreted as local and readily accessible in the vicinity of the site [@pereira_territory_2016-1]. Frequently, the quartz and greywacke assemblages include large amounts of shattered and unknapped chunks or slabs, and have been interpreted in great part as functionally specialised in activities related to the fragmentation of bones in the case of quartz and greywack [@cascalheira_2017], or in grease rendering in the case of quartz [@manneIntensiveSubsistencePractices2012]. Previous studies from the Terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiroProtoSolutreanLithicTechnology2020]. Based on this, despite the smaller amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits. The importance of chalcedony can also be seen in its use to produce formal tool kits during specific technocomplexes at the site, such as a Vale Comprido point, the fossil-director for the Proto-Solutrean in Western Portugal [@belmiro_gravettian-solutrean_2021] or a laurel leaf preform in the Solutrean layers [@gibajabaoProvenienceTechnologyMorphology2013].

@@

While the advantages of employing a macroscopic methodology for analysing the cherts of Vale Boi are notable, a limitation of this approach is its inability to unequivocally characterize and categorize artefacts that exhibit surface alterations typical of an open-air site, including those related to fire, since weathered and altered surface may destroy visible structures and even alter the chemical composition [@delluniversitaDevelopmentMultiparametricCharacterisation2019]. Other studies frequently exclude samples that display intense fire damage or several surface alterations [@delluniversitaDevelopmentMultiparametricCharacterisation2019; @gomezdesolerPanadellaChertMontmaneu2020; @sotoProcurementMobilityLate2016]. For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), and given the inexistence of a reference collection for local altered cherts, they were analysed using the individual sample dataset (Online Resource 1) but grouped in a category of indeterminate cherts (INDET). This category included key alterations such as heat/fire related alterations which clearly impacted the macroscopic characteristics of the samples, fully changing its color, luster and texture, and covering them in crazings (e.g., @fig-terrace-alterations a). Similarly, it also included alteration rinds that extended through most of the sample, impeding the identification of its original color, texture, translucency, inclusions and fossils (e.g., @fig-terrace-alterations b) and the presence of extended pitting. However, whenever surface alterations were present in only a limited area of the sample leaving recognizable macroscopic features and fossils, or when key elements were distinguishable to make a characterization (e.g., reddening in the Type 2 cherts, Online Resource 2, figure s5), the samples were not collapsed into the INDET category.

@@

```{r}
#| label: tbl-type-source
#| tbl-cap: Chert types identified during the analysis, and potential sources. The distance column refers to the distance of the potential source from Vale Boi, as the crow flies. Only the TL samples with possible identified sources were included in the table.

@@

Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiroCreatingFramesReference2023], we interpret Types 2, 3, 4, and 5 to be local, as they are congruent with the described characteristics representative of cherts from the Lower Jurassic found in the known outcrops Cabo S. Vicente, Foz dos Fornos, Ponta dos Altos, Praia do Belixe and Ferrel (Carixian; \<20 km from Vale Boi, as the crow flies), possibly Upper Jurassic Kimmeridgian outcrops of Mareta (\~15 km from Vale Boi, as the crow flies) and Middle Jurassic Malhão formations (80-100 km East of Vale Boi, as the crow flies) of the Algarve basin. Types 2 and 3 are congruent with the frequently available brown/yellow and red nodules and beds (@fig-local-types b-d), ranging from uncommon to very frequent fossils. These can be found in all previously mentioned outcrops. The macroscopic similarities of Types 4 and 5 are more limited (@fig-local-types e-h), as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the inland Lower Jurassic outcrop of Ferrel (@fig-map b). As such, we interpret these cherts as local, but possibly from nodules and inland outcrops that are no longer frequently available or visible in the landscape. This is especially relevant since the inland outcrops of Ferrel are currently mostly destroyed and can be found in poor preservation conditions, due to recent human intervention and construction work, and with small, broken nodules in the surface, frequently altered [@belmiroCreatingFramesReference2023].

@@

Studies of Upper Cretaceous cherts located in the Lisbon area describe cherts with similar macroscopic characteristics (@fig-map a), with micro-cryptocrystalline quartz mudstones with bioclasts such as sponge spicules or bivalve fragments, or wackestone/packstone textures with abundant bioclasts, including ostracods and different types of foraminifers [@jordaoFlintSourcesMobility2022]. These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from Central Portugal hosted at LusoLit. Similar to the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matiasRawMaterialSourcing2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), iron oxides and low porosity (\<1%). From these Cretaceous cherts from Central Portugal, only the dark grey nodules (sample RT231, with petrographic descriptions available in our online research compendium) showed identifiable fossils which were rare ostracods.

@@

The consultation of the lithotheque hosted at UCA allowed us to identify a subset of cherts with similarities to those from Type 7 from the Middle Subbetic region (Betic Systems, south Spain, @fig-map). These cherts are massive, sub-translucent, peloidal/oolitic packstones. Allochems include very frequent peloids and ooids. The peloids are densely arranged in the samples, more or less visible depending on the alterations to the surface or the thickness. The ooids are poorly sorted, with oval or round shapes, and replaced by quartz. Uncommonly the ooids show a preserved nucleus. The presence of bioclasts varies between samples, ranging between common to rare, and when present/identifiable include sponge spicules (common), foraminifera (rare to uncommon), and echinoderm spines (rare). Similar cherts have also been identified and described in previous works, describing grey cherts with peloidal and oolitic facies and occasional foraminifera [@rodriguezPrehistoricFlintExploitations2011]. Despite the non-existence of thin sections for comparison, the macroscopic similarities between the cherts allow us to interpret Type 7 as coming from the Upper Jurassic formations of the Betic Systems, located in the south and southeastern Iberian Peninsula, at least 250 km (as the crow flies) east from Vale Boi (@fig-map).

Type 8 (@fig-distance-types, e-f) corresponds to a massive micro-cryptocrystalline quartz from a marine environment and shelf/platform depositional environment. It represents \~2% of the overall Terrace cherts and \~5% of the chert in the Shelter. Identified and well-preserved fossils are foraminifera from the Pfenderenidae family (@fig-distance-types e, f2), which are larger benthic foraminifera from marine environments [@boudagher-fadelChapterMesozoicLarger2008]. The association of these foraminifera indicates this chert formed from sediments deposited possibly between the Jurassic to the Cretaceous. Macroscopically, this type is characterized by grey and white colours (@fig-distance-types e-f1). The translucency ranges from sub-translucent to opaque. Whenever present, the cortex is rounded and thin (@tbl-cortex_tab). Petrographically, this type is a packstone (mudstone texture close to the edges) with a massive microstructure. It is composed of 97% micro-cryptocrystalline quartz, macrocrystalline quartz, and fibrous chalcedony mostly replacing fossils. Other minerals include uncommon iron oxides and the porosity is of vug type (3%). Fossil content is very frequent (@fig-distance-types f2-f3), albeit mostly unidentifiable and poorly preserved. Identified fossils are rare sponge spicules, rare ostracods with some degree of preservation, common small, unidentified foraminifera (100 μm) of differing preservation degrees, and as mentioned above, foraminifera from the Pfenderenidae family (500 μm). Despite the macroscopic similarities between type 8 and cherts from the Middle Subbetic region chert formations (@fig-map), the geological reference samples showed no macroscopically visible foraminifera. However, our results indicate that despite no foraminifera being visible macroscopically, they are present in thin section. The fact that no foraminifera were identified in the small number of analysed samples from the UCA lithotheque with similarities to Type 8 (n=3) may simply reflect their uncommon visibility, possibly due to patination. Another possible source for Type 8 cherts may be the Upper Cretaceous wackestone/packstone cherts with foraminifers from the Lisbon region [@jordaoFlintSourcesMobility2022]. However, these studies do not specify the existence of Pfenderenidae foraminifers. Given the lack of this specific fossil in the reference samples from the Betic chert \~250 km east and southeast, and the Central Portugal cherts 215-250 km north, the probable source of Type 8 remains unknown.

Type 9 (@fig-distance-types g) corresponds to a banded micro-cryptocrystalline quartz/micrite peloidal packstone. It corresponds to less than \~2% of the overall cherts in the Terrace and Shelter areas. It is opaque, has a heterogeneous structure and is characterized by a variable colour distribution: it shows a horizontal, finely laminated pattern, with grey dark bands intercalated with light grey bands (@fig-distance-types g1). Whenever present, the cortex is rounded and is generally thin (@tbl-cortex_tab). In specific samples, the cortex is present on two parallel planes of the sample (parallel to the laminations), which indicates the chert was originally available in bedded layers. Petrographically this chert is a banded packstone (banding of sedimentary origin; @fig-distance-types g2), composed mostly of micro-cryptocrystalline quartz (48%), micrite (40%), dolomite (10%), and accessory fibrous chalcedony and mica (muscovite). The allochems are peloids, unidentifiable bioclasts, iron oxides, and opaque minerals. Porosity is low (1%). Although its origin is unknown, when visiting the UCA lithotheque, we identified a small type of banded black cherts without noticeable fossils or inclusions. Despite the lack of comparative petrographic data, the macroscopic resemblances raise the possibility that Type 9 belongs to Betic chert formations in Southern Spain, 250 km east (as the crow flies) from the site (@fig-map).

@@

count_plot <- ggplot(count_table, aes(x=Level,y=percent, fill = RMTYPE)) +
  geom_bar(stat = "identity", width = 0.5) +
  facet_grid(.~Technocomplex, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format()) +
  #theme_minimal() + #type of theme
  labs(y = "Count %") +# change labels
  guides(fill=guide_legend(title=NULL)) + # remove legend title
  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
                position=position_stack(vjust=0.5), colour="white", size=3) +
  theme(axis.title.x = element_blank()) +
  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
                              "#996666", "#BDD2B6", "#798777", "#BDC7C9"))

@@
